---
title: "analysis"
author: "Manunpat"
date: '2022-04-30'
output: html_document
editor_options: 
  chunk_output_type: console
---

1. Model without regressors

### 1. Check stationary
The plot and the kpss test show that the time series is not stationary
```{r}
oilwti_tsibble %>% features(log(DCOILWTICO),unitroot_kpss) 
#pvalue (0.01) is very low so we reject the null hypothesis that the time series is stationary
```
### 2. Find a proper diferencing for non-stationary
The unitroot test shows that only one difference is proper to create stationary
```{r, echo = FALSE, message = FALSE, warning=FALSE}
oilwti_tsibble %>% features(log(DCOILWTICO),unitroot_ndiffs) #1
oilwti_tsibble %>% features(difference(log(DCOILWTICO)),unitroot_kpss)
#pvalue is 0.1 which is large enough to not reject the null hypothesis
```
### 3. Examine ACF and PACF to determine model
It seems there are autocorrelation in lag 2-5 
```{r, echo = FALSE, message = FALSE, warning=FALSE}
gg_tsdisplay(oilwti_tsibble,difference(log(DCOILWTICO)), plot_type = "partial")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#arima by hand: ignore seasonality because seasonality is very small
fit <- model(oilwti_tsibble,
              arima310 = (ARIMA(log(DCOILWTICO) ~ pdq(3,1,0))),
              arima013 = (ARIMA(log(DCOILWTICO) ~ pdq(0,1,3))),
              stepwise = (ARIMA(log(DCOILWTICO))),
              search = ARIMA(log(DCOILWTICO), stepwise = FALSE))
report(fit)
#stepwise=search is the best model from AIC, AICc, BIC - ARIMA(2,1,3)(1,0,0)[52]
#stepwise 0.00495   2333. -4652. -4652. -4613. <cpl [54]> <cpl [3]>
fit

#residual check: overall perfect
fit %>% select(stepwise) %>% gg_tsresiduals()
#no autocorrelation

resi_fit <- fit %>% select(stepwise) %>% residuals(type = c("innovation"))
qqPlot(resi_fit$.resid)
#there are a few outliers due to the covid pandamic. Except those periods the residuals are almost normal distributed. 

Box.test(resi_fit$.resid, type = "Ljung")
#pvalue (0.9791) is high so our residual is independently distributed

#forecast
fit %>% forecast(h = 52) %>% autoplot(oilwti_tsibble, level = 0) + coord_cartesian(ylim = c(90, 120))
#+ coord_cartesian(ylim = c(3, 5.5), xlim = c(18000,19500))
```

2. Model with regressors
ARIMA + Regression
AIC check fit in general because AIC exaimine fit model with null model
```{r echo=FALSE, message=FALSE, warning=FALSE}
#2 variables
fit_reg <- model(gdpcpi,
                 mod_gdpcpi = ARIMA(log(DCOILWTICO) ~ logGDP + logCPI),
                 mod_gdp    = ARIMA(log(DCOILWTICO) ~ logGDP),
                 mod_cpi    = ARIMA(log(DCOILWTICO) ~ logCPI))


report(fit_reg) 
ts(fit_reg$mod_gdp)

# model 1 is the best (lowest AIC, AICc, BIC).-> we choose model_gdpcpi for a scenario analysis
#1 mod_gdpcpi 0.00463   2369. -4723. -4722. -4678. <cpl [54]> <cpl [1]>
#2 mod_gdp    0.00488   2319. -4623. -4622. -4578. <cpl [54]> <cpl [2]>
#3 mod_cpi    0.00468   2358. -4701. -4701. -4656. <cpl [54]> <cpl [2]>
#AIC, AICc, BIC improves 1.4%-1.5% so we chose mod_gdpcpi

fit_reg %>% select(mod_gdpcpi) %>% gg_tsresiduals()

resi_fit <- fit_reg %>% select(mod_gdpcpi)  %>% residuals(type = c("innovation"))
qqPlot(resi_fit$.resid)
#there are a few outliers due to the covid pandamic. Except those periods the residuals are almost normal distributed. 

Box.test(resi_fit$.resid, type = "Ljung")

#residuals
#bind_rows('Regrssion Error' = as_tibble(residuals(fit2, type = "regression")),
#          'ARIMA ERROR' = as_tibble(residuals(fit2, type = "innovation")), 
#          .id = "type") %>%
#  ggplot(aes(x = observation_date, y = .resid)) + geom_line() + 
#  facet_grid(vars(type), scales = "free_y") + labs(y="")
```

3. Scenario Analysis
```{r echo=FALSE, message=FALSE, warning=FALSE}

fit_final <- fit_reg %>% select(mod_gdpcpi) 

#scenario 1: normal situation (CPI +2%, US GDP +2.5%)
# Historical US GDP before Covid is around 2.5%
#https://data.worldbank.org/indicator/NY.GDP.MKTP.KD.ZG?locations=US
# The inflation is in line with the target from the Federal Reserves at 2%
future <- new_data(gdpcpi, 52) %>% mutate(logCPI = NA,logGDP = NA)
future$logCPI[1] <- last(gdpcpi$logCPI)
future$logCPI[52] <- last(gdpcpi$logCPI) + log(1.02)
future$logGDP[1] <- last(gdpcpi$logGDP)
future$logGDP[52] <- last(gdpcpi$logGDP) + log(1.025)
future$logCPI <- na.approx(future$logCPI)
future$logGDP <- na.approx(future$logGDP)

forecast(fit_final, new_data = future) %>%
  autoplot(oilwti_tsibble) + 
  geom_line(aes(y = .fitted, color = "Fitted"), data = augment(fit_final))
  labs(y = "Percentage change")  

#scenario 2: Financial crisis (CPI + 0%, US GDP -3%) 
# From the historical US GDP, GDP changes in 2008 and 2020 were -2.5% and -3.4%.
# Assume that CPI is unchanged because CPI normally increases
future_crisis <- new_data(gdpcpi, 52) %>% mutate(logCPI = NA,logGDP = NA)

future_crisis$logCPI[1] <- last(gdpcpi$logCPI)
future_crisis$logCPI[52] <- last(gdpcpi$logCPI) + log(1)
future_crisis$logGDP[1] <- last(gdpcpi$logGDP)
future_crisis$logGDP[52] <- last(gdpcpi$logGDP) + log(1-0.03)
future_crisis$logCPI <- na.approx(future_crisis$logCPI)
future_crisis$logGDP <- na.approx(future_crisis$logGDP)  

forecast(fit_final, new_data = future_crisis) %>%
  autoplot(oilwti_tsibble) + 
  geom_line(aes(y = .fitted, color = "Fitted"), data = augment(fit_final))
  labs(y = "Percentage change")

  
#scenario 3: Ukrain Russia conflict (CPI + 8%, limited impact on US GDP) 
# If the war is contained within the region (Ukraine and Russia. i.e. no war escalation to the other regions), the impact on US economy is limited
# Reference: https://www.bls.gov/cpi/
future_war <- new_data(gdpcpi, 52) %>% mutate(logCPI = NA,logGDP = NA)

future_war$logCPI[1] <- last(gdpcpi$logCPI)
future_war$logCPI[52] <- last(gdpcpi$logCPI) + log(1.08)
future_war$logGDP[1] <- last(gdpcpi$logGDP)
future_war$logGDP[52] <- last(gdpcpi$logGDP) + log(1.025)
future_war$logCPI <- na.approx(future_war$logCPI)
future_war$logGDP <- na.approx(future_war$logGDP)

forecast(fit_final, new_data = future_war) %>%
  autoplot(oilwti_tsibble) + 
  geom_line(aes(y = .fitted, color = "Fitted"), data = augment(fit_final))
  labs(y = "Percentage change")

```  



