---
title: "WTI OIL"
subtitle: "Forecasting"
author: "Jiaming Huang, Manunpat Sirijaturaporn, Ting Yang"
date: "2022/06/01"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts]
    lib_dir: libs
    nature:
      highlightStyle: arta
      highlightLines: true
      countIncrementalSlides: false
      
---


```{r xaringan-logo, echo=FALSE}
xaringanExtra::use_logo(
image_url = "https://wp.unil.ch/bisa/files/2015/10/hec_logo_white.png",
position = xaringanExtra::css_position(top = "0.5em", right = "1em"))
```

# Objective, Questions and Outline
<br>
<span style="font-size:24px;">__Objective__: Make a prediction of oil prices and perform scenario analysis  
<br>
<span style="font-size:24px;">__Questions__: <br>
<span style="font-size:24px;"> 1. Find the best model for oil price prediction <br> 
<span style="font-size:24px;"> 2. Predict oil prices in different economic scenarios <br>
<br>
<span style="font-size:24px;">__Outline__: <br>
>-  <span style="font-size:20px;"> Overview of time series   
   
>- <span style="font-size:20px;"> Model without regressors   

>- <span style="font-size:20px;"> Model with regressors   

>- <span style="font-size:20px;"> Scenario analysis
---
```{r, include=FALSE}
source(here::here("script/setup.R"))
oilwti <- read_excel(here::here("data/DCOILWTICO.xls"))
oilwti$observation_date <- yearweek(oilwti$observation_date)
oilwti_tsibble <- as_tsibble(oilwti)
gdp <- read_excel(here::here("data/GDP.xls"))

gdp_quarterly <- gdp
gdp_quarterly$observation_date <- yearquarter(gdp_quarterly$observation_date)
gdp_quarterly <- as_tsibble(gdp_quarterly)
gdp_quarterly$GDP <- log(gdp_quarterly$GDP)
cpi <- read_excel(here::here("data/USACPIALLMINMEI.xls"))
cpi_monthly <- cpi
cpi_monthly$observation_date <- yearmonth(cpi_monthly$observation_date)
cpi_monthly <- as_tsibble(cpi_monthly)
cpi_monthly$USACPIALLMINMEI <- log(cpi_monthly$USACPIALLMINMEI)
gdp_weekly <- gdp
gdp_weekly$observation_date <- yearweek(gdp_weekly$observation_date)
gdp_weekly$GDP <- log(gdp_weekly$GDP)
gdp_weekly <- as_tsibble(gdp_weekly) 
gdp_weekly <- gdp_weekly %>% fill_gaps() %>% 
  mutate(logGDP = zoo::na.approx(GDP)) %>% 
  select(observation_date, logGDP)
has_gaps(gdp_weekly) #FALSE
cpi_weekly <- cpi
cpi_weekly$observation_date <- yearweek(cpi_weekly$observation_date)
cpi_weekly$USACPIALLMINMEI <- log(cpi_weekly$USACPIALLMINMEI)
cpi_weekly <- as_tsibble(cpi_weekly) 
cpi_weekly <- cpi_weekly %>% fill_gaps() %>% 
  mutate(logCPI = zoo::na.approx(USACPIALLMINMEI)) %>% 
  select(observation_date, logCPI)
has_gaps(cpi_weekly) #FALSE
gdpcpi <- oilwti_tsibble %>% left_join(gdp_weekly) %>% left_join(cpi_weekly)
gdpcpi <- gdpcpi[1:1878,]
```

```{r setup, include=FALSE}
options(
  htmltools.dir.version  = FALSE,
  htmltools.preserve.raw = FALSE # needed for windows
)
```

# Overview of time series

__Several key nodes of the shocks: __  
- Gulf war at Iraq and Kuwait in 1990s
- The economic bubbles and financial crisis in 2000s (~2008)
- The excessive oil supply in 2014
- Covid pandemic in 2020
- Currently, the price recovery due to the lockdown lift and the Ukraine-Russia war

```{r echo=FALSE, message=FALSE, fig.align='center', fig.height=4, fig.width=11}
#ts plotting

p1 <- oilwti_tsibble %>% autoplot()+
  labs(y = "WTI Oil price ($)", title = "Weekly time series of WTI price")

plotly::ggplotly(p1)
```

data updated: 2022-05-14
---
# STL Component

<br>
```{r echo=FALSE, message=FALSE, fig.align='center', fig.width=11, fig.height=5}
dcmp <- oilwti_tsibble %>% model(STL(log(DCOILWTICO)))
components(dcmp) %>% autoplot() + xlab("Year") +ylab("logWTI") 
```

- Trend: It looked stable before 1999 then there had been an upward trend until 2009. After 2009, it is stable again
- Seasonality: There is a yearly seasonality but it has a small impact on this time series
- Remainder: Normally it is volatile around the range of 0.5 to -0.5 but we see the sharp drop in 2020 
---

# Building Model | without regressors
<br>
.pull-left[
```{r, echo=FALSE, message=FALSE, warning=FALSE, out.height=0.7}

# - Kpss test shows that the ts is not stationary  
# - Unit root test shows that only one difference is proper 
# - In ACF and PACF plot, there are autocorrelation in the first 2 or 3 lags   
fit <- model(oilwti_tsibble,
              arima310 = (ARIMA(log(DCOILWTICO) ~ pdq(3,1,0))),
              arima013 = (ARIMA(log(DCOILWTICO) ~ pdq(0,1,3))),
              stepwise = (ARIMA(log(DCOILWTICO))),
              search = ARIMA(log(DCOILWTICO), stepwise = FALSE))
# fit
# 
# fit_desc <- fit %>% pivot_longer(cols = everything(),
#                names_to = "model_type",
#                values_to = "model_specifics_mdl") %>%
#   mutate(model_specifics = format(model_specifics_mdl)) %>%
#   select(-model_specifics_mdl)
```

```{r, include=FALSE}
fit_desc <- fit %>% 
  as_tibble() %>%
  gather() %>%
  mutate(model_desc = print(value)) %>%
  select(key, model_desc) %>%
  set_names("model", "model_desc") 
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=5}
library(kableExtra)
fit_desc%>%
  knitr::kable( "html") %>%
  kable_styling(bootstrap_options = "striped", font_size =12)%>%
  column_spec(1, bold = T, border_right = T, border_left = T, width = "15em") %>%
  column_spec(2, width = "15em")
```
 ]
 .pull-right[
> <span style="color:#03457C ;">Select 1 differencing in ARIMA model and try serveral possibilities.
</span>

]
<br>
- <span style="color:#03457C ;">Here, we can observe auto ARIMA models (stepwise and search) have smaller AICc</span>

```{r, echo=FALSE, message=FALSE, warning=FALSE, out.height= 0.8}
library(DT)
glance.fit <- glance(fit) %>% arrange(AICc) %>% select(.model:BIC)

 DT::datatable(glance.fit, options = list(searching = F, pageLength = 2, height=4),
               caption = "Accuracy of Models") %>%  
  formatRound('sigma2', 2) %>%
  formatRound('AIC', 2)%>%
  formatRound('AICc', 2)%>%
  formatRound('BIC', 2) %>%
  formatRound('log_lik', 2)%>%
  formatStyle(columns = c(1, 2, 3, 4, 5, 6, 7), fontSize = '90%')
```

---
# Building Model | without regressors
> Selected model: auto ARIMA model</span>     

.pull-left[
```{r, echo=FALSE, message=FALSE, warning=FALSE,fig.height=7}
 fit %>% select(stepwise) %>% gg_tsresiduals()

```
]
.pull-right[
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=6}
library(car)
resi_fit <- fit %>% select(stepwise) %>% residuals(type = c("innovation"))
qqPlot(resi_fit$.resid)

```
]
- <span style="color:#03457C ;"> Overall, the residual examination returns good performance. There is no autocorrelation and it is fat-tailed distribution with two large outliers from covid pandemic.
---

# Building Model | without regressors

> Forecasting  

```{r, echo=FALSE, message=FALSE, warning=FALSE,fig.width=11, fig.height=5}
 fit %>% forecast(h = 52) %>% autoplot(oilwti_tsibble, level = 0) +
  labs(y = "log WTI", title = "Forecast of model without regressors")
```

- <span style="color:#03457C ;">Every model returns similar results 
- <span style="color:#03457C ;">The price will increase next year.

---

# Building Model | with regressors
<br>
> Auto ARIMA  + regressors 

<br>
```{r, message=FALSE, warning=FALSE}
fit_reg <- model(gdpcpi,
                 mod_gdpcpi = ARIMA(log(DCOILWTICO) ~ logGDP + logCPI),
                 mod_gdp    = ARIMA(log(DCOILWTICO) ~ logGDP),
                 mod_cpi    = ARIMA(log(DCOILWTICO) ~ logCPI))
```

```{r, include=FALSE}
fit_reg_desc <- fit_reg %>% 
  as_tibble() %>%
  gather() %>%
  mutate(model_desc = print(value)) %>%
  select(key, model_desc) %>%
  set_names("model", "model_desc") 
```

<br>
```{r, echo=FALSE, message=FALSE, warning=FALSE}
fit_reg_desc%>%
  knitr::kable( "html", caption ="Models description") %>%
   kableExtra::kable_styling(bootstrap_options = "striped", font_size = 15)%>%
  column_spec(1, bold = T, border_right = T, border_left = T, width = "30em") %>%
  column_spec(2, width = "80em")
```


---
# Building Model | with regressors

```{r, echo=FALSE, message=FALSE, warning=FALSE, out.height=0.8}
glance.fit_reg <- glance(fit_reg) %>% arrange(AICc) %>% select(.model:BIC)
 DT::datatable(glance.fit_reg, options = list(searching = F, pageLength = 3, height= 3),
               caption = "Accuracy of models with regressors") %>%  
  formatRound('sigma2', 2) %>%
  formatRound('AIC', 2)%>%
  formatRound('AICc', 2)%>%
  formatRound('BIC', 2) %>%
  formatRound('log_lik', 2)%>%
  formatStyle(columns = c(1, 2, 3, 4, 5, 6, 7), fontSize = '80%')
```

<br>
     
.pull-left[
```{r, echo=FALSE, message=FALSE, warning=FALSE}
loggdp <- c(fit_reg[[1]][[1]][["fit"]][["model"]][["coef"]][["logGDP"]], fit_reg[[1]][[1]][["fit"]][["par"]][["std.error"]][["logGDP"]])

logcpi <- c(fit_reg[[1]][[1]][["fit"]][["model"]][["coef"]][["logCPI"]],
                fit_reg[[1]][[1]][["fit"]][["par"]][["std.error"]][["logCPI"]])

mod_gdpcpi <- c("coef", "s.e")
mod_gdpcpi.coef <- data.frame(mod_gdpcpi, loggdp,logcpi)

mod_gdpcpi.coef %>% knitr::kable( "html", caption ="Model with two regressors", 
                                  font_size = 10, digits = 2) %>%
  column_spec(1, bold = T, border_right = T, border_left = T, width = "10em") %>%
  column_spec(2, width = "10em")

```
]

.pull-right[
- <span style="color:#03457C ;">Residuals are similar to the auto ARIMA model   
- <span style="color:#03457C ;">GDP and CPI are positive correlated with oil price     
- <span style="color:#03457C ;">Auto ARIMA plus two regressors is the best model     
]

---
# Scenario Analysis
> Scenario 1: forecasting with normal situation (CPI+2%, US GDP +2.5%)

```{r, echo=FALSE, message=FALSE, warning=FALSE,fig.width=11, fig.height=5}
fit_final <- fit_reg %>% select(mod_gdpcpi) 
future <- new_data(gdpcpi, 52) %>% mutate(logCPI = NA,logGDP = NA)
future$logCPI[1] <- last(gdpcpi$logCPI)
future$logCPI[52] <- last(gdpcpi$logCPI) + log(1.02)
future$logGDP[1] <- last(gdpcpi$logGDP)
future$logGDP[52] <- last(gdpcpi$logGDP) + log(1.025)
future$logCPI <- na.approx(future$logCPI)
future$logGDP <- na.approx(future$logGDP)
forecast(fit_final, new_data = future) %>%
  autoplot(oilwti_tsibble) + 
  geom_line(aes(y = .fitted, color = "Fitted"), data = augment(fit_final))+
  labs(y = "WTI Oil price ($)", title = "Scenario forecasting with normal situation")
```

- <span style="color:#03457C ;">GDP: Historical US GDP before Covid is around 2.5%
- <span style="color:#03457C ;">CPI: The inflation is in line with the target from the Federal Reserves at 2%
<br>
[Data from the world bank](https://data.worldbank.org/indicator/NY.GDP.MKTP.KD.ZG?locations=US)


---
# Scenario Analysis
> Scenario 2: forecasting with financial crisis (CPI+0%, US GDP-3%)

```{r, echo=FALSE, message=FALSE, warning=FALSE,fig.width=11, fig.height=5}
future_crisis <- new_data(gdpcpi, 52) %>% mutate(logCPI = NA,logGDP = NA)
future_crisis$logCPI[1] <- last(gdpcpi$logCPI)
future_crisis$logCPI[52] <- last(gdpcpi$logCPI) + log(1)
future_crisis$logGDP[1] <- last(gdpcpi$logGDP)
future_crisis$logGDP[52] <- last(gdpcpi$logGDP) + log(1-0.03)
future_crisis$logCPI <- na.approx(future_crisis$logCPI)
future_crisis$logGDP <- na.approx(future_crisis$logGDP)  
forecast(fit_final, new_data = future_crisis) %>%
  autoplot(oilwti_tsibble) + 
  geom_line(aes(y = .fitted, color = "Fitted"), data = augment(fit_final)) +
  labs(y = "WTI Oil price ($)", title = "Scenario forecasting with financial crisis")
```

- <span style="color:#03457C ;"> GDP: Historical US GDP changes in 2008 and 2020 were -2.5% and -3.4%, respectively   
- <span style="color:#03457C ;"> CPI: Assuming inflation stalls (CPI is unchanged) 
---
# Scenario Analysis
> Scenario 3: forecasting with Ukarin Russia conflict (CPI+8%, limited impact on US GDP)

```{r, echo=FALSE, message=FALSE, warning=FALSE,fig.width=11, fig.height=5}
future_war <- new_data(gdpcpi, 52) %>% mutate(logCPI = NA,logGDP = NA)
future_war$logCPI[1] <- last(gdpcpi$logCPI)
future_war$logCPI[52] <- last(gdpcpi$logCPI) + log(1.08)
future_war$logGDP[1] <- last(gdpcpi$logGDP)
future_war$logGDP[52] <- last(gdpcpi$logGDP) + log(1.025)
future_war$logCPI <- na.approx(future_war$logCPI)
future_war$logGDP <- na.approx(future_war$logGDP)
forecast(fit_final, new_data = future_war) %>%
  autoplot(oilwti_tsibble) + 
  geom_line(aes(y = .fitted, color = "Fitted"), data = augment(fit_final)) +
  labs(y = "WTI Oil price ($)", title = "Scenario forecasting with Ukarin Russia conflict")
```

- <span style="color:#03457C ;"> GDP: War between Ukarin and Russia has limited impact on the US economy   
- <span style="color:#03457C ;"> CPI: [Data from United States Department of Labor](https://www.bls.gov/cpi/)

---

# Q & A 

<br>
<br>
<br>
<br>
<br>
<span style="font-size:40px;">__Thank you for your attention!__

